Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2020-01-12 rnni_convexity.tex
mv def convexity to introduction

We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between tree $T$ and $R$.

@@

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
We call a set of trees \emph{convex} if every pair of trees from this set is connected by a shortest path that stays within this set.
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
Splits are denoted by $A|B$.
We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant, the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
We will also consider the so-called Cluster Theorem as another weak version of the Split Theorem (see Conjecture~\ref{conjecture:cluster_theorem}).

>>>>

We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between tree $T$ and $R$.
When analysing the geometry of $\rnni$, we will consider a subset of trees and see that it is \emph{convex} in $\rnni$, which means that every pair of trees from this subset is connected by a shortest path that stays within this subset.

@@

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
Splits are denoted by $A|B$.
We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant, the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
We will also consider the so-called Cluster Theorem as another weak version of the Split Theorem (see Conjecture~\ref{conjecture:cluster_theorem}).
